---
title: "Data Structures in R"
author: "Thomas Koentges"
date: "10/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Basics 

R is like a fast calculator

```{r}
232047294879347 * 242424 / 12313 + 2
```

In R `<-` is used to assign a value to a variable. You can also use `=`.

```{r}
test_variable <- 232047294879347 * 242424 / 12313 + 2
test_variable
```

You can use `rm() to remove a variable or object.

```{r}
test_variable
rm(test_variable)
test_variable = 2
test_variable
```

You do not have to remove to update or change the variable and you can change the variable using it's own value.

```{r}
test_variable <- test_variable * test_variable
test_variable
```

Aside from Base R functions, you can include third party functions, objects, and data via libraries
```{r}
# install.packages("tidyverse")
# This without the leading # would be the way to install a library.
# A line with a leading # is ignored by R.
# I have done so, because I do not want to install a package, already installed on my computer.
# You only have to use the install.packages() function once on each new system
# or after upgrading your R version.

# You can also use those #-lines to comment your code
# if you are using RStudio have a look what happens when you
# start a line with # and finish the line with ####
# This feature of RStudio helps you to organise your code.
library(tidyverse)
mpg %>% ggplot(aes(displ,hwy)) + geom_point()
# you do not need to understand the above code yet, just understand that we are using an external library
```
The tidyverse package is a library that contains a lot of useful features and it is the package, I use in almost every R programme.  
You can also generate your own functions:
```{r}
myFunction <- function(x) {
  result <- x + 2
  return(result)
}
myFunction(15)
```

But before we make it anymore complicated than this, let's have a look at the different data types and structures R has to offer.

# 2. R's Data Types 

The datatype of each variable is inferred:  
Logical is either `TRUE` or `FALSE`.
```{r}
variable <- FALSE 
class(variable)
```

Numeric is a number that can include fractual parts, i.e. 2.34.

```{r}
variable <- 2.34 
class(variable)
```

Integer is a whole number.

```{r}
variable <- 2L 
variable
class(variable)
```

##### Sidetrack
Integers are cheaper than Numerics:
```{r}
object.size(c(1L,2L,3L,4L,5L)) < object.size(c(1,2,3,4,5))
```
Unless they are empty:
```{r}
object.size(numeric()) == object.size(integer())
```
Then they are both 48bytes:
```{r}
object.size(numeric())
```
##### Back on track

You can also express complex numbers.
```
variable <- 3 + 2i
variable
class(variable)
```
And plot them:
```{r}
plot(variable)
```

Strings are called characters in R.
```{r}
variable <- "hello world"
variable
class(variable)
```

The last data type is `raw`. This refers to a byte stream.
```{r}
variable <- charToRaw(variable)
variable
class(variable)
```

Raw objects are small. Empty raw objects are also 48bytes though:

```{r}
object.size(raw())
```

The types you will use most are characters, numeric, and logical.

# 3. Data Structures

Technically everything that you have produced so far are vectors. Vectors can be combined using the `c()` function. Vectors can only contain one type of data. They will throw an error or normalise if you try to force two different types together.

```{r}
VectorA <- 2
VectorB <- c(1L,2L)
c(VectorA,VectorB)
class(c(VectorA,VectorB))
class(c(VectorB))
```

A list is an R-object which can contain many different types, such as vectors, functions and even another list inside it.

```{r}
list(VectorA, VectorB, list(VectorA, VectorB), mean)
```

A matrix is a two-dimensional rectangular data set of ONE data type. It can be created using a vector input to the matrix function. This is how you create a matrix:
```{r}
M <- matrix( c('a','a','b','c','b','a'), nrow = 2, ncol = 3, byrow = TRUE)
M
```

If you want to know who the `matrix()` function works, you can ask R:
```{r}
?matrix
```

Let's have a look at the result, if we use the same input vector, but a different method of filling the matrix:
```{r}
M <- matrix(c('a','a','b','c','b','a'), nrow = 2, ncol = 3, byrow = FALSE)
M
```

Do you see how it has changed? 

A factor stores the vector and all its unique values.

```{r}
# Create a vector.
apples <- c('iOS','iPhone','MBP','iPhone','MBP')
# Create a factor object.
factor_apples <- factor(apples, levels = c("iOS", "iPhone", "MBP", "Mac"), ordered = T)
factor(apples)

# Have a look
apples
factor(apples)
factor_apples
nlevels(factor_apples)
factor_apples[4]
as.character(factor_apples[4])
```

And finally, Data Frames are mixed data presented in tabular form. 

```{r}
# Create the data frame.
data_types <- data.frame(
  types = c("logical", "numerical","integer", "complex", "character", "raw"), 
  usage_probability = c(1, 1, 1, 0.1, 1, 0.5), 
  got_it = c(rep("yes", 5), "no"),
  answer = rep(42L, 6)
)
boxplot(usage_probability ~ types, data_types)
```

The `tidyverse` has also a modernised version of the dataframe called tibble:

```{r}
library(tidyverse)
data_types_tibble <- as_tibble(data_types)

# Let's have a look
data_types
data_types_tibble
```
You might already see the advantages of the tibble.

# 4. Other data Structures often used in programming

## Strings and Arrays
Strings in R are called characters and they behave differently than in other programming languages:
```{r}
newString <- "test"
length(newString)
nchar(newString)
```

One dimensional arrays of simple types in R are best expressed as vectors. And so, if another developer talks about "arrays", the R programmer should translate that to "vectors". There is actually a data structure called "array" in R, but this is mainly reserved for multi-dimensionional arrays:

```{r}
intArray <- c(7L,8L,9L,10L)
length(intArray)
intArray 

## sidetrack, you do not need to remember this
intRealArray <- array(c(7L,8L,9L,10L),dim = c(2,2))
intRealArray
```

## Hash table
R does not have hash tables. Instead, R uses something like it that is called environment or it uses lists. That said ther is a `hashmap` library.

## Linked Lists
Linked list are directly implemented in R using list:

```{r}
lst <- list() # creates an empty (length zero) list
lst
lst[[1]] <- 1 # automagically extends the lst
lst[[2]] <- 2 # ditto
lst

lst <- vector("list", 5)
lst
lst <- list(1, 2, 3, 4, 5)
lst
lst <- lst[-2]
lst
```

## Trees

In R trees are just a list of lists. As an aside, R sometimes is not very memory efficient (if you do not watch out) and it likes copying (a lot). And so, if you want to store the data as a more complex graph, you need to use a graph db.

Here is an example of a simple tree:

```{r}
tree <- list(list(1, 2), list(3, list(4, 5)))

# left child: list(1, 2)
tree[[1]]

# right child
tree[[2]]

# left child of right child:list(4, 5)
tree[[2]][[1]]
```

That's all. Feel free to go over this page again at a later stage. Those are some solid mechanics for more complex problems.
